# Reproducible Research: Peer Assessment 1

This project examines data from a personal activity monitoring device.
The data covers the months of October and November, 2012, and it consists
of the number of steps taken in each 5-minute interval in those months.


## Loading and preprocessing the data

The data consists of three columns:

* Steps: #steps taken

* Date: date of observation

* Interval: minute of day corresponding to start of 5-minute interval of 
  observation

```{r load-data}
obs.data <- read.csv("activity.csv", header=TRUE, 
                     colClasses=c("integer", "Date", "integer"))
summary(obs.data)
```
The time for the interval data is stored in the format HHMM, which is very 
misleading when plotted as an integer. So we convert it into the format mmmm,
which is the number of minutes since midnight.
```{r convert-data}
obs.data$interval <- 60 * (obs.data$interval %/% 100) + (obs.data$interval %% 100)
summary(obs.data)
```

## What is mean total number of steps taken per day?

The first question to be answered is the mean total number of steps taken
per day.  To find this, first compute the total number of steps taken
each day, then average the results.

```{r aggregate-by-day}
obs.by.date <- aggregate(steps ~ date, obs.data, sum)
summary(obs.by.date)
```

As you can see, the **mean** total number of steps taken per day is
`r sprintf("%.2f", mean(obs.by.date$steps))`, and the 
**median** is `r sprintf("%d", median(obs.by.date$steps))`.  The following
figure gives a better overview of the distribution.
```{r total-number-of-steps-per-day}
hist(obs.by.date$steps, breaks=20, col="red",
     main="Total Number of Steps Taken per Day",
     xlab="Steps Taken", ylab="Frequency")
```

## What is the average daily activity pattern?

The next question concerns the average daily activity pattern, or how
the activities varies by time.  This is best seen in a line graph,
after averaging the data by time of day.

```{r plot-aggregate-by-time-of-day}
obs.by.interval <- aggregate(steps ~ interval, obs.data, mean)
plot(obs.by.interval, type="l", col="red",
     main="Steps Taken on an Average Day",
     xlab="Minute of Day",
     ylab="# of Steps")

```

From the graph, it is clear that **peak activity** occurs around the 500th
minute, or 8am.  Analytically, we can also find the actual interval with
the peak activity and convert it to a time as follows.

```{r max-steps-by-time-of-day}
idx <- which.max(obs.by.interval$steps)
minute <- obs.by.interval[idx, "interval"]
paste ((minute %/% 60), (minute %% 60), sep=":")
```

## Imputing missing values

Unfortunately, the data has a lot of missing values.  As the following
table shows, there are `r sum(is.na(obs.data$steps))` **missing values**
in the steps column.  The other columns do not contain any missing values.

```{r data-summary}
summary(obs.data)
```

There are different ways of eliminating these missing values, such as
filling with the previous known value or using an average.  For this 
study, it makes sense to fill with the median value based on the time
interval. There is great variability based on the time of day, so that
definitely should be taken into account. Moreover, the median is more
robust in the prescence of outliers, such as a day walking at the park 
or relaxing at the beach.  The effect is easily seen in a histogram 
as follows.  The big change is that there now a significant number of
days when between 1,000 and 2,000 steps were taken.

```{r fill-in-missing-values}
obs.by.interval.median <- aggregate(steps ~ interval, obs.data, median)
obs.data.fill <- merge(x = obs.data, 
                       y = obs.by.interval.median, 
                       by = "interval")
obs.data.fill$steps <- obs.data.fill$steps.x
obs.data.fill$steps.x <- NULL
nans <- which(is.na(obs.data.fill$steps))
obs.data.fill$steps[nans] <- obs.data.fill$steps.y[nans]
obs.data.fill$steps.y <- NULL
summary(obs.data.fill)
obs.by.date.fill <- aggregate(steps ~ date, obs.data.fill, sum)
hist(obs.by.date.fill$steps, breaks=20, col="red",
     main="Total Number of Steps Taken per Day (After Filling)",
     xlab="Steps Taken", ylab="Frequency")

```

More analytically, the data can be summarized as follows.  

```{r fill-in-missing-values-summary}
summary(obs.by.date.fill)
```

Notice
there are no missing values, and that the **mean** and **median** 
number of steps per day are now 
`r sprintf("%.2f", mean(obs.by.date.fill$steps))`
and `r sprintf("%d", median(obs.by.date.fill$steps))`, respectively.
Compared to the original values of 
`r sprintf("%.2f", mean(obs.by.date$steps))` and 
`r sprintf("%d", median(obs.by.date$steps))`,
this confirms that the effect of removing the NA values 
is to illustrate the days in which between 1,000 and 2,000
steps were taken, which are considerably below the original
mean and median.

## Are there differences in activity patterns between weekdays and weekends?

The final analysis concerns the effect of weekends on activity
levels.  First, each datapoint is classified as either corresponding
to a weekday or weekend.  Then, the data is separated using this
classification, and each subset of the data is separately averaged.
The result, which follows, is a graph clearly showing the different
activity patterns on weekends.

```{r activity-on-weekends}
obs.data.fill$weekday <- weekdays(obs.data.fill$date)
obs.data.fill$daytype <- "weekday"
obs.data.fill$daytype[obs.data.fill$weekday == 'Saturday'] <- "weekend"
obs.data.fill$daytype[obs.data.fill$weekday == 'Sunday']   <- "weekend"
obs.data.fill$daytype = factor(obs.data.fill$daytype)
obs.data.fill$weekday <- NULL

obs.data.fill.weekday <- subset(obs.data.fill, daytype == "weekday")
obs.data.fill.weekend <- subset(obs.data.fill, daytype == "weekend")

obs.by.interval.weekday <- aggregate(steps ~ interval, 
                                     obs.data.fill.weekday, 
                                     mean)
obs.by.interval.weekday$daytype <- "weekday"
obs.by.interval.weekend <- aggregate(steps ~ interval, 
                                     obs.data.fill.weekend, 
                                     mean)
obs.by.interval.weekend$daytype <- "weekend"
obs.by.interval.combined <- rbind(obs.by.interval.weekday, 
                                 obs.by.interval.weekend)
obs.by.interval.combined$daytype <-
  factor(obs.by.interval.combined$daytype)


library(ggplot2)
ggplot(obs.by.interval.combined, 
       aes(x=interval, y=steps)) + 
  labs(title="Steps Taken on an Average Day",
       x="Minute of Day",
       y="# of Steps") +
  geom_line() +
  facet_grid(daytype ~ .)
```

As the graph clearly shows, less activity occurs in the early morning
of weekends, as the subjects are more likely to sleep in.  However,
the activity overall is higher during the days on weekends than on
weekdays, perhaps reflecting a sedentary workplace.
